---
title: "Assignment 2: Coding Basics"
author: "Autumn Dunn"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 
```{r}
#### 1 #### 
#generate sequence from 1 to 100 increase by 4
seq1 <- seq(1,100,4)
print(seq1)
```

2. Compute the mean and median of this sequence. 
```{r}
#### 2 ####
#mean of sequence
mean(seq1)

#median of sequence
median(seq1)
```

3. Ask R to determine whether the mean is greater than the median. 
```{r}
#### 3 #### 
#is the mean greater than the median?
#is 49 > 49
if (mean(seq1) > median(seq1)) {
  print("This is true!")
} else {
  print("This is false :(")
}
```

4. Insert comments in your code to describe what you are doing.


## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.
```{r}
#### 5 ####
#Name students - Character vector
students <- c("Sam", "Max", "Liz", "Jen")
print(students)

#Generate random test scores - numeric vector
testscores <- c(sample(1:100, 4))
print(testscores)

#Determine if the student will pass - logical vector
passing <- testscores >50
print(passing)
```

6. Label each vector with a comment on what type of vector it is.
```{r}
#### 6 ####
### vectors are labeled with a comment above ###
#students = character vector
#testscores = numeric vector
#passing = logical vector
```

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}
#### 7 ####
#Create the teacher's grade book
GradeBook <- data.frame(students, testscores, passing)
print(GradeBook)
```

8. Label the columns of your data frame with informative titles.
```{r}
#### 8 ####
#Assign Names to each column
x_name <- "Student"
y_name <- "Test Score"
z_name <- "Passing"

#label the gradebook
names(GradeBook) <- c(x_name, y_name, z_name)
print(GradeBook)
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix is a collection of elements of the same data type (numeric, character, or logical). 
Our data frame is a MIXTURE of data types, with numeric data (testscores), character data (students), AND logical data (passing). 
Maxtrix = all columns with same type of data
Data Frame = columns with different types of data

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#### if and else statement function ####
#create function that determines if x is a passing grade
function1 <-function(x) {
if (x > 49) {
  ("This student is passing")
} else {
  ("This student is failing")
}}

##  Attempt 1 ##
funpassingONE_attempt1<-function1(testscores)
print(funpassingONE_attempt1)
#this did not work

##  Attempt 2 ##
#make the test scores an array
scoresoftest<-cbind(testscores)
print(scoresoftest)
#apply the function to each individual value
funpassingONE_attempt2<-apply(scoresoftest,c(1,2), function1)
print(funpassingONE_attempt2)
#this worked

#Create grade book 2
GradeBook2 <- data.frame(students, testscores, funpassingONE_attempt2)
print(GradeBook2)


#### ifelse function ####
#create function that determines if y is a passing grade
function2<-function(y) {
ifelse (y > 49, "This student is passing","This student is failing") }

##  Attempt 1 ##
funpassingTWO_attempt1<-function2(testscores)
print(funpassingTWO_attempt1)
#this worked

#Create grade book 3
GradeBook3 <- data.frame(students, testscores, funpassingTWO_attempt1)
print(GradeBook3)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked and 'if' and 'else' did not work. In 'if' and 'else', the condition is only applied to the first element. ifelse is applied to all elements. To make the 'if' and 'else' function work, you must apply the function to each individual test score. Whereas, the ifelse will apply it to each individual test score.
