---
title: "Assignment 7: Time Series Analysis"
author: "Autumn Dunn"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on time series analysis. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A07_TimeSeries.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 16 at 11:59 pm.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1
getwd()
#install packages and add to library
if(!require(pacman)) install.packages("pacman")
pacman::p_load(ggpubr, ggplot2, readxl, tidyverse, dplyr, lubridate, trend, zoo, readr, Kendall)

#Graph Theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        legend.position = "top") 
theme_set(mytheme)

#import data
EPAair_2019 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv")
EPAair_2018 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv")
EPAair_2017 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv")
EPAair_2016 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv")
EPAair_2015 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv")
EPAair_2014 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv")
EPAair_2013 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv")
EPAair_2012 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv")
EPAair_2011 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv")
EPAair_2010 <- read_csv("./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv")

GaringerOzone<- bind_rows(EPAair_2019, EPAair_2018, EPAair_2017, EPAair_2016, EPAair_2015, EPAair_2014, EPAair_2013, EPAair_2012, EPAair_2011, EPAair_2010)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3.
GaringerOzone$Date <- mdy(GaringerOzone$Date)
class(GaringerOzone$Date)

# 4.
wrangle_GaringerOzone <- GaringerOzone %>% 
  select(Date, `Daily Max 8-hour Ozone Concentration`, DAILY_AQI_VALUE)

# 5.
Days <- as.data.frame( x = seq(as.Date("2010-01-01"), as.Date("2019-12-31"), "day")) %>% 
  rename("Date" = "seq(as.Date(\"2010-01-01\"), as.Date(\"2019-12-31\"), \"day\")")

# 6.
GaringerOzone_final <- left_join(x = Days, y = wrangle_GaringerOzone) %>% 
  rename("Ozone.ppm" = "Daily Max 8-hour Ozone Concentration",
         "AQI" = "DAILY_AQI_VALUE") 
```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7.
ggplot(GaringerOzone_final, aes(x = Date, y = Ozone.ppm)) +
  geom_line() +
  geom_smooth(method = lm) +
  ylab("Daily Max Ozone Concentration (ppm)") 
```

>Answer: From 2010 to 2020, ozone concentration has a decreasing trend. There appears to be some seasonal patterns when looking at the none smoothed line, where ozone decreases in winter and increases in summer.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8.
# Adding new column with no missing obs
GaringerOzone_clean <- 
  GaringerOzone_final %>% 
  mutate( Ozone.ppm_clean = zoo::na.approx(Ozone.ppm) )

#Compare clean and NA ozone
ggplot(GaringerOzone_clean) +
  geom_line(aes(x = Date, y = Ozone.ppm_clean), color = "red") +
  geom_line(aes(x = Date, y = Ozone.ppm), color = "black") +
  ylab("Daily Max Ozone Concentration (ppm)") 

#Just clean Ozone data
ggplot(GaringerOzone_clean) +
  geom_line(aes(x = Date, y = Ozone.ppm_clean)) +
  ylab("Daily Max Ozone Concentration (ppm)") 
```

> Answer: Ozone levels fluctuate. Using a piecewise constant would assume the values would be the same as the previous or following day, but there were no instances of ozone levels being the same two days in a row, so piecewise is not the best way to estimate missing values. Quadratic function would not make sense estimating ozone levels either. Linear interpolation estimates a middle point between one observation and the next, which makes the most sense for ozone estimations.

9. Create a new data frame called `GraingerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9.
GaringerOzone.mean <- GaringerOzone_clean %>% 
  mutate(Year = year(Date),
         Month = month(Date)) %>% 
  group_by(Year, Month) %>% 
  summarise(Mean_Ozone = mean(Ozone.ppm_clean)) %>% 
  ungroup() %>% 
  mutate(Day = 1)

GaringerOzone.mean$Date<-as.Date(with(GaringerOzone.mean,paste(Year,Month,Day,sep="-")),"%Y-%m-%d") 

GaringerOzone.monthly <- GaringerOzone.mean %>% 
                         select(Date, Mean_Ozone) 
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10.
# Generate time series
D_month <- month(first(GaringerOzone_clean$Date))
D_year <- year(first(GaringerOzone_clean$Date))

GaringerOzone.daily.ts <- ts(GaringerOzone_clean$Ozone.ppm_clean,
                   start=c(D_year,D_month),
                   frequency=365) 
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$Mean_Ozone,
                   start=c(D_year,D_month),
                   frequency=12) 
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11.
#decompose
Day_Ozone_decomp <- stl(GaringerOzone.daily.ts, s.window = "periodic")
plot(Day_Ozone_decomp)

Month_Ozone_decomp <- stl(GaringerOzone.monthly.ts,s.window = "periodic")
plot(Month_Ozone_decomp)
```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12.
# Run SMK test
Month_Ozone_trend1 <- Kendall::SeasonalMannKendall(Month_Ozone_ts)

# Inspect results
Month_Ozone_trend1
summary(Month_Ozone_trend1)
```

> Answer: Seasonal Mann-Kendall is the most appropriate because we do have a distinct seasonal pattern that differs from the trend and data patterns. 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13.
#Visualization
Month_Ozone_plot <-
ggplot(GaringerOzone.monthly, aes(x = Date, y = Mean_Ozone)) +
  geom_point() +
  geom_line() +
  ylab("Monthly Mean Ozone Concentration (ppm)") +
  geom_smooth( method = lm )+ 
  scale_x_date(date_labels = "%Y", breaks = "year")
print(Month_Ozone_plot)
```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: Have ozone concentrations changed over the 2010s at this station? Based on the graph and the Seasonal Mann Kendall test, date and ozone concentratation has a significant downward monotonic trend (tau = -0.143, p-value < 0.5). Ozone concentrations have changes over the 2010s at this station, ozone has been decreasing.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15.
  # extract the components and turn them into data frames
GaringerOzone_Components <- as.data.frame(Month_Ozone_decomp$time.series[,1:3])

GaringerOzone_Components <- mutate(GaringerOzone_Components,
                            Observed = GaringerOzone.monthly$Mean_Ozone,     
                            Date = GaringerOzone.monthly$Date)

Nonseasonal_ozone <- GaringerOzone.monthly.ts - GaringerOzone_Components$seasonal
plot(Nonseasonal_ozone)
plot(GaringerOzone.monthly.ts)

#16.
Month_Ozone_trend2 <- Kendall::MannKendall(Nonseasonal_ozone)
Month_Ozone_trend2
summary(Month_Ozone_trend2)
```

> Answer: The non-seasonal ozone mann kendall test has a smaller p-value (0.00754) than seasonal ozone mann kendall test (0.0467). Non-seasonal ozone had a smaller tau value (-0.165) than seasonal (-0.143). This indicates that without seasonal influences, ozone has a very signficiant downward monotonic trend.
